To Batch or Not to Batch: Sequential vs. Batched Testing Strategies in the ED

### 2023-10-06 manuscript/manuscript.tex
aking updated to manuscript

\usepackage{booktabs}
\usepackage{siunitx}
\usepackage{tablefootnote}
\usepackage{longtable}
\usepackage{threeparttable}
\usepackage{natbib}

@@

\hypertarget{sec:1}{%
\section{Introduction}\label{sec:1}}

@@

We start by evaluating the reduced-form effects of ED provider batch
ordering tendency on downstream patient outcomes and turnaround time. We
find that practice variation as captured by physician batch-ordering
tendency has large and significant consequences. Being treated by a
provider in the top decile of the leniency distribution, compared to
being treated by someone in the bottom decile {[}INSERT RESULTS{]}.

Because of the institutional features of the ED, our research design
closely approximates an RCT that assigns patients to batch-ordering or
sequential-ordering arm. In the ED, patients have no discretion over
choosing providers, and in our specific ED, physicians have discretion
over choosing patients, alleviating major selection issues present in
other health care settings. Furthermore, physicians exhibit wide
variation in practice behavior in batch ordering, even within the same
hospital, while following the same guidelines. Finally,
patient-physician interactions in the ED are typically well documented,
short, and one-off, constraining physician decision-making to a more
limited, better-observed choice set than present in settings such as
specialty or primary care.

In sum, exploiting practice variation in ED settings shuts down other
(but not all) potential channels besides test batching that are present
in other settings, determine length of stay, and impact patient
outcomes. This approach allows us to move closer to identifying the
causal impact of batch-ordering diagnostic tests on patient outcomes and
hospital effeciency. It is important to note that this paper studies the
impact of an batch-ordering through a batching decision requiring
clinical judgment (within practice norms) rather than through specific
hospital policies, differences in adherence to clinical practice
guidelines, or substandard care.

The remainder of this paper is structured as follows. The next section
describes the data source and outlines our baseline sample. The
empirical strategy and its accompanying identifying assumptions are laid
out in Section II. Section III presents the results. Section IV draws
implications for batch-ordering policies. The last section concludes.

\hypertarget{sec:2}{%
\section{Data and Definitions}\label{sec:2}}

@@

We identified three distinct forms of test batching:

\begin{itemize}
\item
  Lab/Image Batch: The first test ordered for a patient was either an
  imaging or lab test (CT scan, X-ray, Ultrasound, Lab), followed by an
  additional imaging or lab test order within 5 minutes.
\item
  Image/Image Batch: The first test ordered for a patient was an imaging
  test (CT scan, X-ray, Ultrasound), followed by an additional imaging
  test order within 5 minutes.
\item
  Any Batch: This refers to instances where either a Lab/Image Batch or
  an Image/Image Batch occurred.
\end{itemize}

We conducted a sensitivity analysis for this definition, documented in
Appendix 2. Table 1 highlights the considerable variation in patient
complexity between those who had their tests batch ordered versus those
where tests were ordered sequentially.

@@

Table 1 provides a comprehensive breakdown of the patient encounters
involving diagnostic tests ordered either individually (Sequential) or
as part of a group (Batch) during an ED encounter from our baseline
sample. Around 43\% of patient encounters involved diagnostic tests
ordered as part of a batch, the majority of which (60\%) included two
tests.

Differences were noted between the two cohorts. Patients who had batched
tests ordered tended to have a longer ED length of stay, a lower
Emergency Severity Index (ESI), and were generally older than those who
had their tests ordered individually. Patients in the batch order group
were also more likely to present with symptoms at triage, including
tachycardia, tachypnea, fever, and hypotension. A deeper dive into the
ESI data reveals a higher proportion of patients with more urgent
conditions (ESI 1 and 2) in the batched group than in the sequential
group. This finding might be indicative of the complexity and severity
of the conditions of patients who necessitate batch testing.

Age distributions between the two groups demonstrate a greater
proportion of patients aged 65 and above in the batch group, suggesting
that older patients are more likely to have multiple tests ordered at
once, possibly due to the presence of multiple comorbidities. The
batched group showed a longer average ED length of stay compared to the
sequential group. This extended stay may be reflective of the additional
time required to conduct and evaluate the results of multiple tests.
Symptoms presentation at triage also differed between the two groups,
with the batched group being more likely to present with tachypnea,
fever, and hypotension. However, tachycardia was found to be similar
across both groups.

\begin{table}[htbp]
\centering
\tiny
\caption{Summary Statistics for Sequential vs. Batched Test Groups}
\begin{tabular}{lcccc}
\hline
Variable & \multicolumn{1}{c}{Overall Visits, N = 41,197\textsuperscript{1}} & \multicolumn{1}{c}{Sequential Arm, N = 23,4331} & \multicolumn{1}{c}{Batching Arm, N = 17,7641} & \multicolumn{1}{c}{p-value\textsuperscript{2}} \\
\hline
\textbf{ESI} & & & & $<$0.001 \\
\hspace{0.5cm} 1 & 474 (1.2\%) & 130 (0.6\%) & 344 (1.9\%) & \\
\hspace{0.5cm} 2 & 13,911 (34\%) & 6,502 (28\%) & 7,409 (42\%) & \\
\hspace{0.5cm} 3 & 23,655 (57\%) & 14,350 (61\%) & 9,305 (52\%) & \\
\hspace{0.5cm} 4 & 3,123 (7.6\%) & 2,422 (10\%) & 701 (3.9\%) & \\
\hspace{0.5cm} 5 & 34 ($<$0.1\%) & 29 (0.1\%) & 5 ($<$0.1\%) & \\
\hline
\textbf{Age} & & & & $<$0.001 \\
\hspace{0.5cm} $<$20 & 790 (1.9\%) & 605 (2.6\%) & 185 (1.0\%) & \\
\hspace{0.5cm} 20-45 & 9,412 (23\%) & 6,242 (27\%) & 3,170 (18\%) & \\
\hspace{0.5cm} 45-65 & 12,826 (31\%) & 7,393 (32\%) & 5,433 (31\%) & \\
\hspace{0.5cm} 65+ & 18,169 (44\%) & 9,193 (39\%) & 8,976 (51\%) & \\
\hline
\textbf{ED LOS} & & & & $<$0.001 \\
\hspace{0.5cm} Mean (SD) & 270 (382) & 258 (177) & 286 (545) & \\
\hline
\textbf{72hr Return} & 1,482 (3.6\%) & 957 (4.1\%) & 525 (3.0\%) & $<$0.001 \\
\hline
\textbf{Number of Tests} & & & & $<$0.001 \\
\hspace{0.5cm} 1 & 12,961 (31\%) & 11,937 (51\%) & 1,024 (5.8\%) & \\
\hspace{0.5cm} 2 & 19,112 (46\%) & 8,457 (36\%) & 10,655 (60\%) & \\
\hspace{0.5cm} 3 & 7,688 (19\%) & 2,581 (11\%) & 5,107 (29\%) & \\
\hspace{0.5cm} 4 & 1,342 (3.3\%) & 432 (1.8\%) & 910 (5.1\%) & \\
\hspace{0.5cm} 5 & 94 (0.2\%) & 26 (0.1\%) & 68 (0.4\%) & \\
\hline
\textbf{Disposition} & & & & $<$0.001 \\
\hspace{0.5cm} Discharge & 25,302 (61\%) & 15,922 (68\%) & 9,380 (53\%) & \\
\hspace{0.5cm} Admit & 9,097 (22\%) & 4,209 (18\%) & 4,888 (28\%) & \\
\hspace{0.5cm} Other & 6,798 (17\%) & 3,302 (14\%) & 3,496 (20\%) & \\
\hline
\textbf{Tachycardic} & 8,115 (20\%) & 4,562 (19\%) & 3,553 (20\%) & 0.18 \\
\hline
\textbf{Tachypneic} & 3,822 (9.3\%) & 1,669 (7.1\%) & 2,153 (12\%) & $<$0.001 \\
\hline
\textbf{Febrile} & 969 (2.4\%) & 353 (1.5\%) & 616 (3.5\%) & $<$0.001 \\
\hline
\textbf{Hypotensive} & 651 (1.6\%) & 305 (1.3\%) & 346 (1.9\%) & $<$0.001 \\
\hline
\textbf{Gender} & & & & $<$0.001 \\
\hspace{0.5cm} Female & 22,105 (54\%) & 12,986 (55\%) & 9,119 (51\%) & \\
\hspace{0.5cm} Male & 19,092 (46\%) & 10,447 (45\%) & 8,645 (49\%) & \\
\hline
\textbf{Race} & & & & $<$0.001 \\
\hspace{0.5cm} White & 36,554 (89\%) & 20,665 (88\%) & 15,889 (89\%) & \\
\hspace{0.5cm} Black & 1,682 (4.1\%) & 974 (4.2\%) & 708 (4.0\%) & \\
\hspace{0.5cm} Asian & 1,217 (3.0\%) & 710 (3.0\%) & 507 (2.9\%) & \\
\hspace{0.5cm} Other & 737 (1.8\%) & 449 (1.9\%) & 288 (1.6\%) & \\
\hspace{0.5cm} Native & 538 (1.3\%) & 340 (1.5\%) & 198 (1.1\%) & \\
\hspace{0.5cm} Unknown & 469 (1.1\%) & 295 (1.3\%) & 174 (1.0\%) & \\
\hline
\hline
\multicolumn{5}{l}{\textsuperscript{1} n (\%)} \\
\multicolumn{5}{l}{\textsuperscript{2} Pearson's Chi-squared test; Welch Two Sample t-test} \\
\end{tabular}
\end{table}

\begin{figure}
\centering
\includegraphics[width=\textwidth,height=0.3\textheight]{../Manuscript/figures/fig1.png}
\caption{Frequency of Sample Cases and Batches}
\end{figure}

@@

We document that the Mayo Clinic ED physicians exhibit wide, systematic
variation in their propensity to batch order diagnostic tests. Figure 2
graphs the histogram of the leniency measure along the x-axis and the
left y-axis. A local-linear regression of the fitted probability of
batched ordering on tendency after residualizing is overlaid and
displayed on the right y-axis. Table 2 presents the same ``first stage''
in a regression table: being assigned to a 10 pp more lenient physician
is associated with a 17 pp increase in the likelihood of being
prescribed an opioid in the ED. The F-statistic is 25 when all controls
and fixed effects are included. The coefficient is greater than one
because all emergency visits are used to construct the leniency
instrument, while the first stage is calculated using the baseline
sample only, which excludes the rarely prescribed diagnoses.

@@

Where \(Y_i\) represents our main outcomes of interest: length of stay,
72 hour readmission, and resource utilization.

\begin{table}[htbp]
\centering
\caption{First Stage Regression Results}
\begin{tabular}{lccc}
\toprule
 & Model 1 & Model 2 & Model 3 \\
\midrule
Lab + Image Batching Tendency & 0.99*** & 1.02*** & 1.02*** \\
 & (0.01) & (0.01) & (0.01) \\
\\
F statistic (proj model) & 13206.68 & 6702.01 & 6655.86  \\
F (proj model): p-value & $<0.001$ & $<0.001$ & $<0.001$ \\
\midrule
Image + Image Batching Tendency & 0.92*** & 0.94*** & 0.94*** \\
 & (0.02) & (0.04) & (0.04) \\
\\
F statistic (proj model) & 1645.35 & 577.97 & 577.09  \\
F (proj model): p-value & $<0.001$ & $<0.001$ & $<0.001$ \\
\midrule
\midrule
Num. obs. & 41197 & 41197 & 41197 \\
Seasonality and shift fixed effects? & Yes & Yes & Yes \\
Chief Complaint? & No & Yes & Yes \\
Patient observables? & No & No & Yes \\
\bottomrule
\end{tabular}
\begin{tablenotes}
\small
\item \textit{Notes:}  Estimates of the first stage for the baseline sample described in the text. Seasonality shift fixed effects include Year-Month and Hospital-Day of week-Hour of day fixed effects. Chief complaint comes from the cleaned complaint that the patient came in with at the initial encounter. Patient observables include sex dummy, race/ethnicity, and age bins. Column 3 corresponds to the baseline controls. Robust standard errors are clustered at the physician level.
\item $*** p < 0.001$, $** p < 0.01$, $* p < 0.05$.
\end{tablenotes}
\end{table}

and \(X_i\) is the same as in the reduced-form approach. \(Batched_i\)
variable suffers from potential endogeneity concerns. For example,
injury severity may be unobserved and correlated with need to run
multiple tests, which in turn also affects length of stay. Hence, we
instrument \(Batched_i\) with the assigned physician \(j\)'s underlying
tendency to batch, \(Tendency_{i,j}^{phys}\) We cluster robust standard
errors at the physician level to account for the assignment process of
patients to physicians.

\hypertarget{identifying-assumptions}{%
\subsubsection{Identifying Assumptions}\label{identifying-assumptions}}

The reduced-form approach delivers an unbiased estimate of the causal
effect of being treated by a higher tendency to batch physician, if
assignment of patients to ED physicians is random, conditional on
seasonality, shift, and diagnosis (``conditional independence''). One
can test for whether batch-likely physicians are routinely assigned
patients with particular characteristics. Figure 4 shows evidence of
balance on observables. The left panel is the output of a regression of
Prescribed on observables including patient demographics, priority
groups, and ED diagnoses. Unsurprisingly, many observables predict
length of stay and 72 hour return. The right panel displays the
coefficients from a regression of physician tendency instrument on the
same set of observables. With so many coefficients, we fail to reject
that the instrument violates conditional independence (the joint F-stat
drops by two orders of magnitude, from 280 to 2.4). In fact, some of the
variables that are significant in the second column go in the opposite
direction of the decision to batch

The residualization in equation (1) controls for more controls than
required to achieve quasi-random assignment; they are included for
statistical precision in measuring physician tendency to batch. In
Appendix Figure {[}XXX{]}, residualizing for only seasonality and shift
(left panel) and also diagnosis (right panel) in equation (1). This
lends support to our assumption that assignment in the ED is nearly
random at the seasonality, shift level.

Our instrumental variable approach, which aims to recover the causal
effect of having diagnostic tests batch ordered, relies on three
additional assumptions: relevance, exclusion, and monotonicity. We
reported a strong first stage (i.e., relevance) at the end of the
previous Section. The exclusion restriction requires that the instrument
must influence the outcome of interest only through its effect on test
batching. This is perhaps our strongest assumption and is at its core,
untestable. However, several features of the ED setting suggest that
such violation may likely only have a small impact and may be less
concerning than in other health care settings. First, unlike in primary
care settings, where the patient and primary care provider have many
repeat encounters, the scope of what the emergency physician can do to
impact medium-term outcomes is limited and well-observed by the
researcher. Second, any violation of the exclusion restriction needs to
directly affect the specific outcome of interest. The channel by which
ED physicians can influence length of stay relative outcomes is likely
through testing and diagnosis. Nevertheless, we take this assumption
seriously and perform a placebo check in Section IIIB as well as various
robustness checks in Section IIIE.

\hypertarget{results}{%
\subsection{Results}\label{results}}

\hypertarget{reduced-form-results}{%
\subsubsection{Reduced-Form Results}\label{reduced-form-results}}

@@

@@

\begin{table}[ht]
\centering
\caption{Regression Results by Chief Complaint} 
\begin{tabular}{llrr}
  \hline
 & CHIEF\_COMPLAINT & Coefficient\_Starred & P\_value \\ 
  \hline
1 & Extremity Complaints & -0.72** & 0.03 \\ 
  2 & Upper Respiratory Symptoms & 0.426** & 0.02 \\ 
  3 & Abnormal Test Results & -0.067 & 0.84 \\ 
  4 & Falls, Motor Vehicle Crashes, Assaults, and Trauma & -0.423* & 0.10 \\ 
  5 & Shortness of Breath & -0.065 & 0.58 \\ 
  6 & Gastrointestinal Issues & 0.071 & 0.67 \\ 
  7 & Chest Pain & 0.107 & 0.68 \\ 
  8 & Dizziness/Lightheadedness/Syncope & -0.163 & 0.32 \\ 
  9 & Neurological Issue & -0.353 & 0.25 \\ 
  10 & Fatigue and Weakness & -0.112 & 0.49 \\ 
  11 & Back or Flank Pain & -0.602** & 0.02 \\ 
  12 & Abdominal Complaints & -0.105 & 0.13 \\ 
  13 & Fevers, Sweats or Chills & 0.01 & 0.95 \\ 
  14 & Skin Complaints & 1.442 & 0.35 \\ 
   \hline
\end{tabular}
\end{table}

\begin{table}[ht]
\centering
\caption{Regression Results by Chief Complaint (2)} 
\begin{tabular}{llrr}
  \hline
 & CHIEF\_COMPLAINT & Coefficient\_Starred & P\_value \\ 
  \hline
1 & Extremity Complaints & 0.176 & 0.50 \\ 
  2 & Upper Respiratory Symptoms & 0.332*** & 0.00 \\ 
  3 & Abnormal Test Results & 0.695 & 0.14 \\ 
  4 & Falls, Motor Vehicle Crashes, Assaults, and Trauma & -0.247 & 0.20 \\ 
  5 & Shortness of Breath & -0.253** & 0.01 \\ 
  6 & Gastrointestinal Issues & 0.421*** & 0.00 \\ 
  7 & Chest Pain & -0.531*** & 0.00 \\ 
  8 & Dizziness/Lightheadedness/Syncope & 0.075 & 0.60 \\ 
  9 & Neurological Issue & -0.305 & 0.15 \\ 
  10 & Fatigue and Weakness & -0.125 & 0.40 \\ 
  11 & Back or Flank Pain & 0.288 & 0.13 \\ 
  12 & Abdominal Complaints & -0.023 & 0.66 \\ 
  13 & Fevers, Sweats or Chills & -0.044 & 0.71 \\ 
  14 & Skin Complaints & 0.328 & 0.35 \\ 
   \hline
\end{tabular}
\end{table}

\begin{table}[ht]
\centering
\caption{First Stage over Subsamples} 
\begin{tabular}{llrr}
  \hline
 & CHIEF\_COMPLAINT & Coefficient\_Starred & P\_value \\ 
  \hline
1 & Extremity Complaints & 0.761*** & 0.00 \\ 
  2 & Upper Respiratory Symptoms & 1.406*** & 0.00 \\ 
  3 & Abnormal Test Results & 1.036*** & 0.00 \\ 
  4 & Falls, Motor Vehicle Crashes, Assaults, and Trauma & 0.831*** & 0.00 \\ 
  5 & Shortness of Breath & 1.258*** & 0.00 \\ 
  6 & Gastrointestinal Issues & 1.043*** & 0.00 \\ 
  7 & Chest Pain & 0.866*** & 0.00 \\ 
  8 & Dizziness/Lightheadedness/Syncope & 1.668*** & 0.00 \\ 
  9 & Neurological Issue & 0.723*** & 0.00 \\ 
  10 & Fatigue and Weakness & 1.651*** & 0.00 \\ 
  11 & Back or Flank Pain & 0.752*** & 0.01 \\ 
  12 & Abdominal Complaints & 1.109*** & 0.00 \\ 
  13 & Fevers, Sweats or Chills & 1.23*** & 0.00 \\ 
  14 & Skin Complaints & 0.448* & 0.07 \\ 
   \hline
\end{tabular}
\end{table}

>>>>

\usepackage{booktabs}
\usepackage{tabularx}
\usepackage{siunitx}
\usepackage{tablefootnote}
\usepackage{longtable}
\usepackage{threeparttable}
\usepackage{natbib}

@@

\hypertarget{sec:I}{%
\section{Introduction}\label{sec:I}}

@@

We start by evaluating the reduced-form effects of ED provider
batch-ordering tendency on downstream patient outcomes and turnaround
time. We find that practice variation as captured by physician
batch-ordering tendency has large and significant consequences. Being
treated by a provider in the top decile of the tendency distribution,
compared to being treated by someone in the bottom decile {[}INSERT
RESULTS{]}.

Because of the institutional features of the ED, our research design
closely approximates an RCT that assigns patients to batch-ordering or
sequential-ordering arm. In the ED, patients have no discretion over
choosing providers, and in our specific ED, physicians have discretion
over choosing patients, alleviating major selection issues present in
other health care settings. Furthermore, physicians exhibit wide
variation in practice behavior in batch-ordering, even within the same
hospital, while following the same guidelines. Finally,
patient-physician interactions in the ED are typically well documented,
short, and one-off, constraining physician decision-making to a more
limited, better-observed choice set than present in settings such as
specialty or primary care.

In sum, exploiting practice variation in ED settings shuts down other
(but not all) potential channels besides test batching that are present
in other settings, determine length of stay, and impact patient
outcomes. This approach allows us to move closer to identifying the
causal impact of batch-ordering diagnostic tests on patient outcomes and
resource utilization. It is important to note that this paper studies
the impact of batch-ordering through a batching decision requiring
clinical judgment (within practice norms) rather than through specific
hospital policies, differences in adherence to clinical practice
guidelines, or substandard care.

The remainder of this paper is structured as follows. The next section
describes the data source and outlines our baseline sample. The
empirical strategy and its accompanying identifying assumptions are laid
out in Section III. Section IV presents the results. Section V draws
implications for batch-ordering policies. The last section concludes.

\hypertarget{sec:II}{%
\section{Data and Definitions}\label{sec:II}}

@@

In the context of diagnostic test ordering, we defined batching as
placing multiple test orders within 5 minutes of one another.
Sensitivity analyses on this cutoff point showed that our results are
robust to this definition. Our study differentiated between two distinct
batching strategies:

\begin{itemize}
\item Lab + Image Batch: The batch order for a patient consisted of two or more distinct diagnostic tests where at least one imaging (contrasted CT scan, non-contrasted CT scan, X-ray, ultrasound) and one lab test were placed.

\item Image + Image Batch: The batch order for a patient consisted of two or more distinct imaging (contrasted CT scan, non-contrasted CT scan, X-ray, ultrasound) diagnostic tests.
\end{itemize}

The distinction is necessary because laboratory tests may be processed
in parallel with imaging tests, which have a much more limited effect on
patient waiting time. However, a batch of multiple imaging tests
requires that the patient is present for both tests. Thus, waiting times
are much more likely to be impacted. Therefore, the specific tests that
make up a batch may provide insight into how the physician trades off
between waiting times and uncertainty, given the various complexity
levels of patient encounters. Figure 1 shows the rate at which each
batching type occurs for the top ten most frequently occurring chief
complaint fields.

\begin{figure}[h]
  \centering
  \includegraphics[width=1\textwidth]{../manuscript/figures/frequency_batches.png}
  \tiny
  \caption{Figure 1 displays two statistics for the ten most common major chief complaint categories observed in our ED visit sample: the unadjusted batching rate for lab-image batching and image-image batching.}
\end{figure}

@@

Table 2 provides a comprehensive breakdown of the patient encounters
involving diagnostic tests ordered individually in a batched or
sequential way during an ED encounter from our baseline sample. About
39\% of patient encounters involved diagnostic tests ordered as part of
a batch, most of which (64\%) included two tests.

Notable distinctions emerge when comparing the two cohorts. Patients
subjected to batched test ordering exhibited a notably prolonged stay in
the emergency department (ED) (p\textless0.001) and were more frequently
assigned a lower Emergency Severity Index (ESI). Furthermore, this group
skewed towards older demographics, with half of them falling into the
65+ age category. Patients in the batch-order group also displayed a
higher likelihood of presenting symptoms at triage, including
tachycardia (p\textless0.001), tachypnea (p\textless0.001), fever
(p\textless0.001), and hypotension (p\textless0.001). Delving deeper
into the ESI data unveils a higher concentration of patients with more
critical conditions (ESI 1 and 2) within the batched group, in contrast
to the sequential group (45.1\% with ESI of 1 or 2 in the batched group
compared to 27.5\% with ESI of 1 or 2 in the sequential group). This
finding suggests that factors related to patient complexity and severity
(Saghafian et al., 2014) significantly influence physicians' decisions
to opt for batch-ordering of diagnostic tests. However, it is important
to note that our investigation revealed substantial variation among
physicians in their choices between batching and sequencing for a given
patient.

\begin{table}
\centering
\caption{Summary Statistics}
\begin{tabular}{p{3cm} *{4}{p{2.5cm}}p{2.5cm}p{2.5cm}p{2.5cm}p{2.5cm}p{2.5cm}p{2.5cm}p{2.5cm}p{2.5cm}}
\toprule
Variable & Overall Visits & Sequentially Arm & Batch Arm & $p$-value \\
&  ($N = 41,846$) & ($N = 25,354$)  & ($N = 16,492$) & \\
\midrule
ESI & & & & $<0.001$ \\
1 & 476 (1.1\%) & 136 (0.5\%) & 340 (2.1\%) & \\
2 & 14,020 (34\%) & 6,865 (27\%) & 7,155 (43\%) & \\
3 & 24,006 (57\%) & 15,517 (61\%) & 8,489 (52\%) & \\
4 & 3,283 (7.9\%) & 2,787 (11\%) & 496 (3.0\%) & \\
5 & 34 (<0.1\%) & 33 (0.1\%) & 1 (<0.1\%) & \\
(Missing) & 27 & 16 & 11 & \\
\midrule
Age & & & & $<0.001$ \\
<20 & 1,413 (3.4\%) & 1,154 (4.6\%) & 259 (1.6\%) & \\
20-45 & 9,419 (23\%) & 6,538 (26\%) & 2,881 (17\%) & \\
45-65 & 12,829 (31\%) & 7,764 (31\%) & 5,065 (31\%) & \\
65+ & 18,185 (43\%) & 9,898 (39\%) & 8,287 (50\%) & \\
\midrule
ED LOS (min) & & & & $<0.001$ \\
Mean (SD) & 269 (379) & 255 (174) & 290 (564) & \\
\midrule
Disposition & & & & $<0.001$ \\
Admit & 9,124 (22\%) & 4,286 (17\%) & 4,838 (29\%) & \\
Discharge & 25,861 (62\%) & 17,596 (69\%) & 8,265 (50\%) & \\
Other & 6,861 (16\%) & 3,472 (14\%) & 3,389 (21\%) & \\
\midrule
Tachycardic & 8,328 (20\%) & 4,853 (19\%) & 3,475 (21\%) & $<0.001$ \\
Tachypneic & 3,953 (9.4\%) & 1,818 (7.2\%) & 2,135 (13\%) & $<0.001$ \\
Febrile & 1,013 (2.4\%) & 386 (1.5\%) & 627 (3.8\%) & $<0.001$ \\
Hypotensive & 69 (1.6\%) & 327 (1.3\%) & 342 (2.1\%) & $<0.001$ \\
\midrule
Number of Tests & & & & $<0.001$ \\
1 & 13,406 (32\%) & 13,406 (53\%) & 0 (0\%) & \\
2 & 19,291 (46\%) & 8,782 (35\%) & 10,509 (64\%) & \\
3 & 7,707 (18\%) & 2,689 (11\%) & 5,018 (30\%) & \\
4 & 1,348 (3.2\%) & 451 (1.8\%) & 897 (5.4\%) & \\
5 & 94 (0.2\%) & 26 (0.1\%) & 68 (0.4\%) & \\
\midrule
Gender & & & & $<0.001$ \\
Female & 22,441 (54\%) & 14,053 (55\%) & 8,388 (51\%) & \\
Male & 19,405 (46\%) & 11,301 (45\%) & 8,104 (49\%) & \\
\midrule
Race & & & & $0.002$ \\
White & 37,070 (89\%) & 22,355 (88\%) & 14,715 (89\%) & \\
Black & 1,727 (4.1\%) & 1,057 (4.2\%) & 670 (4.1\%) & \\
Asian & 1,260 (3.0\%) & 775 (3.1\%) & 485 (2.9\%) & \\
Native & 546 (1.3\%) & 354 (1.4\%) & 192 (1.2\%) & \\
Other & 763 (1.8\%) & 493 (1.9\%) & 270 (1.6\%) & \\
Unknown & 480 (1.1\%) & 320 (1.3\%) & 160 (1.0\%) & \\
\bottomrule
\multicolumn{5}{l}{\textsuperscript{1} n (\%)} \\
\multicolumn{5}{l}{\textsuperscript{2} Pearson's Chi-squared test; Welch Two Sample t-test} \\
\end{tabular}
\end{table}

@@

We document that the Mayo Clinic ED physicians exhibit wide, systematic
variation in their propensity to batch order diagnostic tests. Figure 2
graphs the histogram of batch-ordering frequency by physician for
populat chief complaints, highlighting that the variation in batching
differs systematically. Table 3 presents the ``first stage'' in a
regression table: being assigned to a 10 pp higher tendency physician is
associated with a 10 pp increase in the likelihood of having tests
batch-ordered in the ED. The F-statistic is 51 when all controls and
fixed effects are included.

\begin{table}[htbp]
\centering
\caption{First Stage Regression Results}
\begin{threeparttable}
\begin{tabularx}{0.85\textwidth}{@{}Xccc@{}}
\toprule
 & Model 1 & Model 2 & Model 3 \\
\midrule
Batching Tendency & 0.99*** & 0.99*** & 0.99*** \\
 & (0.01) & (0.01) & (0.01) \\
\midrule
F statistic (full model)&7.32&52.03&50.93\\  
F (full model): p-value& $<0.001$ & $<0.001$ & $<0.001$ \\
\midrule
Num. obs. & 43300 & 43300 & 43300 \\
Seasonality and shift fixed effects? & Yes & Yes & Yes \\
Chief Complaint? & No & Yes & Yes \\
Patient observables? & No & No & Yes \\
\bottomrule
\end{tabularx}
\begin{tablenotes}
\tiny
\item Estimates of the first stage for the baseline sample described in the text. Seasonality shift fixed effects include Year-Month and Hospital-Day of week-Hour of day fixed effects. Chief complaint comes from the cleaned complaint that the patient came in with at the initial encounter. Patient observables include sex dummy, race/ethnicity, and age bins. Column 3 corresponds to the baseline controls. Robust standard errors are clustered at the physician level.
\item $*** p < 0.001$, $** p < 0.01$, $* p < 0.05$.
\end{tablenotes}\tiny
\end{threeparttable}
\label{table:regression}
\end{table}

@@

Where \(Y_i\) represents our main outcomes of interest: length of stay,
72 hour readmission, and resource utilization. and \(X_i\) is the same
as in the reduced-form approach. \(Batched_i\) variable suffers from
potential endogeneity concerns. For example, injury severity may be
unobserved and correlated with need to run multiple tests, which in turn
also affects length of stay. Hence, we instrument \(Batched_i\) with the
assigned physician \(j\)'s underlying tendency to batch,
\(Tendency_{i,j}^{phys}\) We cluster robust standard errors at the
physician level to account for the assignment process of patients to
physicians.

\hypertarget{identifying-assumptions}{%
\subsection{Identifying Assumptions}\label{identifying-assumptions}}

The reduced-form approach delivers an unbiased estimate of the causal
effect of being treated by a higher tendency to batch physician, since
assignment of patients to ED physicians is random, conditional on
seasonality and shift (``conditional independence''). The
residualization in equation (1) controls for more controls than required
to achieve quasi-random assignment; they are included for statistical
precision in measuring physician tendency to batch.

Our instrumental variable approach, which aims to recover the causal
effect of having diagnostic tests batch ordered, relies on three
additional assumptions: relevance, exclusion, and monotonicity. We
reported a strong first stage (i.e., relevance) at the end of the
previous Section. The exclusion restriction requires that the instrument
must influence the outcome of interest only through its effect on test
batching. This is perhaps our strongest assumption and is at its core,
untestable. However, several features of the ED setting suggest that
such violation may likely only have a small impact and may be less
concerning than in other health care settings. First, unlike in primary
care settings, where the patient and primary care provider have many
repeat encounters, the scope of what the emergency physician can do to
impact medium-term outcomes is limited and well-observed by the
researcher. Second, any violation of the exclusion restriction needs to
directly affect the specific outcome of interest. The channel by which
ED physicians can influence length of stay relative outcomes is likely
through testing and diagnosis. Nevertheless, we take this assumption
seriously and perform a placebo check in Section 4.2 as well as various
robustness checks in Section 4.4.

Finally, the monotonicity assumption is necessary for interpreting the
coefficient estimates obtained from the IV approach as Local Average
Treatment Effects (LATEs) if there are heterogeneous treatment effects.
It requires that any patient who is (not) batched by a sequencer
(batcher) would also (not) be batched by a batcher (sequencer)
physician. The literature leveraging the judges design typically
performs two informal tests for its implications. The first one provides
that the first stage should be weakly positive for all subsamples
(Dobbie, Goldin, and Yang 2018). The second implication asserts that the
instrument constructed by leaving out a particular subsample has
predictive power over that same left-out subsample (Bhuller et
al.~2020). Appendix Table 2 presents both of these tests in the two
columns for various subsamples of interest. Finally, we check whether
our main results hold using differential, mutually exclusive leniency
measures (i.e., by complaint category) in Section 4.4.

\hypertarget{sec:4}{%
\section{Results}\label{sec:4}}

\hypertarget{reduced-form-results}{%
\subsection{Reduced-Form Results}\label{reduced-form-results}}

@@

\hypertarget{placebo-check}{%
\subsection{Placebo Check}\label{placebo-check}}

\hypertarget{instrumental-variables-results}{%
\subsection{Instrumental Variables
Results}\label{instrumental-variables-results}}

\hypertarget{robustness}{%
\subsection{Robustness}\label{robustness}}

@@
